---
title: "Results SEIR model"
author: "Eva Santermans"
date: '`r format(Sys.time(), ''%d %B, %Y'')`'
output: html_document
---

```{r,echo=FALSE,message=FALSE}
########################################################
# Modelling the Ebola outbreak in West Africa 
# SEIR model & MCMC estimation
########################################################
# Author: Eva Santermans
# Last update: July 17, 2015
########################################################

## SELECT WORKING DIRECTORY
wd = "C:/Eva/CenStat/Projects/Ebola/Programs - Output/Test/"
setwd(wd)

## SELECT REGION
GEOSEL = "Forecariah"

```


```{r,include=FALSE}
########################################################################################
## LIBRARIES

library(knitr) 
library(markdown)
library(xlsx)
library(MASS)
Sys.setlocale("LC_TIME", "C");
```

```{r,echo=FALSE}
########################################################################################
## CONSTANTS

## Constants
precision=0.1 # precision ODE (days)
pwc.int=21    # period piece-wise constant (days)
```

```{r,include=FALSE}
########################################################################################
## FUNCTIONS

source("data.R")
source("core.R")
```

```{r,echo=FALSE}
########################################################################################
## INPUT 

## Reference data 
index1.cum=diff(rep(cumsum(!is.na(cumcases)),each=1/precision)) # create an index vector to show what period cumulative cases refer to
index2.cum=diff(rep(cumsum(!is.na(deaths)),each=1/precision)) # create an index vector to show what period cumulative deaths refer to
obs1=c(cumcases[!is.na(cumcases)][1],newcases[!is.na(newcases)])
obs2=c(deaths[!is.na(deaths)][1],newdeaths[!is.na(newdeaths)])

## Data frame
J = 5 + n.breaks
mon.names = c("LP","Overdisp1","Overdisp2","E(0)","Proportion fatal","Underreporting","R0",as.parm.names(list(r=rep(0,n.breaks-1))))
parm.names = as.parm.names(list(beta=rep(0,J)))
MyData = list(J=J, mon.names=mon.names, parm.names=parm.names,y=c(obs1,obs2))
```


```{r,include=FALSE}
########################################################################################
## ESTIMATION


## Settings
## --------

set.seed(666)

## Run
## ---

Initial.Values = NULL
res <- LaplacesDemon(Model, Data=MyData, Initial.Values,
                     Iterations=2500000, Status=43898, Thinning=500,
                     Algorithm="AMM", Specs=list(Adaptive=25, B=NULL, n=525000,
                                               Periodicity=898, w=0.05))

## Save and load most recent model
## -------------------------------

name = paste("SEIR_",GEOSEL,".Rdata",sep="")
save(res, file = name)
```

# OUTPUT 
```{r,include=FALSE}
########################################################################################
## CHECKING OUTPUT

## Recommended burn-in for the thinned samples
BurnIn = res$Rec.BurnIn.Thinned
BurnIn

```

```{r,include=FALSE}
## Trace plots
plot(res,BurnIn=1,MyData,PDF=F,Parms=c(0))
```

```{r,include=FALSE}
## Posterior distribution states
post = res$Monitor[-c(1:BurnIn),-1]
E = matrix(NA,dim(post)[1],length(time)); Id = matrix(NA,dim(post)[1],length(time));
Inew = matrix(NA,dim(post)[1],length(time)); M = matrix(NA,dim(post)[1],length(time));
for (i in 1:dim(post)[1]){
  ode=ODEfuncpwc(post[i,],n.cutoffs)
  E[i,] = ode$E; Id[i,] = ode$Id; Inew[i,] = ode$Inew; M[i,] = ode$M
}
```

## Observed and estimated number of cases and deaths. 
```{r,echo=FALSE, fig.width=7, fig.height=7}
rho1 = post[,5];
rho2 = post[,5];
alpha = 1/7.5
preinfperiod = 9.4


sel1 = (length(seq(as.Date("2014-05-23"),min(dates1,na.rm=T),1))-1)*10
sel2 = (length(seq(as.Date("2014-05-23"),min(dates2,na.rm=T),1))-1)*10
time1 = time[1:(length(time)-sel1)]
time2 = time[1:(length(time)-sel2)]

med.Inew = apply(1/preinfperiod*rho1*E,2,median); 
med.M = apply(rho2*alpha*Id,2,median);
med.Inew.cum = apply(rho1*Inew,2,median);
med.M.cum = apply(rho2*M,2,median);
q1.Inew = apply(1/preinfperiod*rho1*E,2,quantile,0.025,na.rm=T); 
q1.M = apply(rho2*alpha*Id,2,quantile,0.025,na.rm=T);
q1.Inew.cum = apply(rho1*Inew,2,quantile,0.025,na.rm=T);
q1.M.cum = apply(rho2*M,2,quantile,0.025,na.rm=T);
q3.Inew = apply(1/preinfperiod*rho1*E,2,quantile,0.975,na.rm=T); 
q3.M = apply(rho2*alpha*Id,2,quantile,0.975,na.rm=T);
q3.Inew.cum = apply(rho1*Inew,2,quantile,0.975,na.rm=T);
q3.M.cum = apply(rho2*M,2,quantile,0.975,na.rm=T);

## plot
par(mfrow=c(2,2),oma = c(0, 0, 2, 0)) 

plot(min(dates1,na.rm=T)+time1,med.Inew[-c(1:sel1)],type="l",xlab="time (in days)",xlim=c(min(dates),max(c(dates1,dates2)+21,na.rm=T)),ylab="cases",ylim=c(0,max(newcases[!is.na(newcases)]/tmp1)*1.2),lwd=2,col="#0571B0")
lines(min(dates1,na.rm=T)+time1,q3.Inew[-c(1:sel1)],type="l",lty=2,lwd=2,col="#0571B0")
lines(min(dates1,na.rm=T)+time1,q1.Inew[-c(1:sel1)],type="l",lty=2,lwd=2,col="#0571B0")
points(dates[!is.na(newcases)]-round(tmp1/2),newcases[!is.na(newcases)]/tmp1,col="black",lwd=1,pch=16)

plot(min(dates2,na.rm=T)+time2,med.M[-c(1:sel2)],type="l",xlab="time (in days)",xlim=c(min(dates),max(c(dates1,dates2)+21,na.rm=T)),ylab="deaths",ylim=c(0,max(newdeaths[!is.na(newdeaths)]/tmp2)*1.2),lwd=2,col="#0571B0")
lines(min(dates2,na.rm=T)+time2,q1.M[-c(1:sel2)],type="l",lty=2,lwd=2,col="#0571B0")
lines(min(dates2,na.rm=T)+time2,q3.M[-c(1:sel2)],type="l",lty=2,lwd=2,col="#0571B0")
points(dates[!is.na(newdeaths)]-round(tmp2/2),newdeaths[!is.na(newdeaths)]/tmp2,col="black",lwd=1,pch=16)

plot(min(dates1,na.rm=T)+time1,med.Inew.cum[-c(1:sel1)],type="l",xlab="time (in days)",xlim=c(min(dates),max(c(dates1,dates2)+21,na.rm=T)),ylab="cumulative cases",ylim=c(0,max(q3.Inew.cum[1:(length(time)-560)],na.rm=T)),lwd=2,col="#0571B0")
lines(min(dates1,na.rm=T)+time1,q1.Inew.cum[-c(1:sel1)],type="l",lty=2,lwd=2,col="#0571B0")
lines(min(dates1,na.rm=T)+time1,q3.Inew.cum[-c(1:sel1)],type="l",lty=2,lwd=2,col="#0571B0")
points(dates[!is.na(cumcases)],cumcases[!is.na(cumcases)],col="black",lwd=1,pch=16)

plot(min(dates2,na.rm=T)+time2,med.M.cum[-c(1:sel2)],type="l",xlab="time (in days)",xlim=c(min(dates),max(c(dates1,dates2)+21,na.rm=T)),ylab="cumulative deaths",ylim=c(0,max(q3.M.cum[1:(length(time)-560)],na.rm=T)),lwd=2,col="#0571B0")
lines(min(dates2,na.rm=T)+time2,q1.M.cum[-c(1:sel2)],type="l",lty=2,lwd=2,col="#0571B0")
lines(min(dates2,na.rm=T)+time2,q3.M.cum[-c(1:sel2)],type="l",lty=2,lwd=2,col="#0571B0")
points(dates[!is.na(deaths)],deaths[!is.na(deaths)],col="black",lwd=1,pch=16)

mtext(GEOSEL, outer = TRUE, cex = 1.5, line=-1)
```
Observed (black) and estimated (blue) number of new cases (top left), new deaths (top right), cumulative cases (bottom left) and cumulative deaths (bottom right) per district. Dashed lines are 95% credible intervals. 

## Estimated effective reproduction number over time. 
```{r,echo=FALSE}
Rvec = post[,6:J]
R = NULL
R = cbind(R,Rvec[,1])

for (i in 2:dim(Rvec)[2]){
  R = cbind(R,apply(Rvec[,1:i],1,sum))
}

med = apply(R,2,median); 
q1 = apply(R,2,quantile,0.025,na.rm=T); 
q3 = apply(R,2,quantile,0.975,na.rm=T); 

med.R = NULL; q1.R = NULL; q3.R = NULL
for (i in 1:(n.breaks)){
  med.R = c(med.R,rep(med[i],n.cutoffs[i]))
  q1.R = c(q1.R,rep(q1[i],n.cutoffs[i]))
  q3.R = c(q3.R,rep(q3[i],n.cutoffs[i]))
}

## plot
par(mfrow=c(1,1),oma = c(0, 0, 2, 0)) #
plot(min(dates1,na.rm=T)+time1,med.R[-c(1:sel1)],col="black",type="l",xlab="time (in days)",ylab=expression(paste(R[e],"(t)")),xlim=c(min(dates),max(c(dates1,dates2)+21,na.rm=T)),ylim=c(0,5),lwd=2)

lines(min(dates1,na.rm=T)+time1,q1.R[-c(1:sel1)],col="black",type="l",lty=2,lwd=2)
lines(min(dates1,na.rm=T)+time1,q3.R[-c(1:sel1)],col="black",type="l",lty=2,lwd=2)
abline(h=1,col="red")

mtext(GEOSEL, outer = TRUE, cex = 1.5, line=-1)
```
Estimated reproduction number per district with 95% posterior intervals. The threshold value of 1 is indicated by a red horizontal line. 

## Parameter estimates with 95% posterior confidence intervals. 
```{r,echo=FALSE}
res$Summary2[(J+3):(2*J+2),c(5:7)]
```
