```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Modelling the Ebola outbreak in West Africa anno 2014
# Date: July 10, 2015
########################################################
library(knitr) 
library(markdown)
library(xlsx)
```

---
title: "Results growth rate model"
author: "Christel Faes"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
  
  # Ebola Working Document: SPATIAL ANALYSIS
  ## Christel Faes
  ## 29 Oct 2014
  
  This document contains the R-code for the analysis of the Ebola outbreak data (case notifications) using linear and non-linear spatial growth models.

### 1) Data section

```{r,echo=FALSE,warning=FALSE,message=FALSE}
##setwd("~/Dropbox/Christel/ONDERZOEK/DataSets/ebola")

# Load R libraries
#-----------------
library(maptools); library(maps); library(RColorBrewer);
library(shapefiles); 
library(INLA);library(MASS)
library(survey)
library(RColorBrewer)
library(classInt)
library(spdep)
options(java.parameters = "-Xmx1000m") 
library(xlsx)
```

#### Reading case notifications per area
```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Read in data
#df=read.xlsx("~/Dropbox/Christel/Onderzoek/datasets/Ebola/Data/EVD_dbase_Reprod_Number_estimation_Macro_20150630.xlsm",sheetIndex=1,startRow=2)

df=read.table("~/R/Rprojects/Ebola model/data.csv",header=T,sep=',')
df<-data

# Drop missing data
df<-df[!is.na(df$Week),]
df<-df[(df$Week!="1900-00"),]
#df<-df[(df$Week!="2015-06"),]
#df<-df[(df$Week!="2015-07"),]

df$week<-(as.numeric(substr(df$Week,3,4)) -14)*52 + as.numeric(substr(df$Week,6,7))
df$week<-df$week-min(df$week)

df$CountW<-as.numeric(as.character(df$CountW))

# Drop missing data
df<-df[!is.na(df$CountW),]
df<-df[!is.na(df$GEO_ID),]
df<-df[,c(1,2,3,5,6,9,11,16)]

# The next areas are not being considered in the analysis
df<-df[df$Outside.West.Africa=="NO",]

df<-df[df$PARENT_COUNTRY_NAM!="Nigeria",]
df<-df[df$PARENT_COUNTRY_NAM!="Senegal",]
df<-df[df$PARENT_COUNTRY_NAM!="Mali",]

# Correction in geographical areas  / areas not in shape file
df$GEO_NAME[df$GEO_NAME == "bo"] <- "Bo"

df<-df[df$GEO_ID != "GNG240735",]
df<-df[df$GEO_ID != "GNG240736",]
df<-df[df$GEO_ID != "GNG240734",]
df<-df[df$GEO_ID != "GNG240711",]
df<-df[df$GEO_ID != "GNG240733",]

# Order data
df<-df[order(df$week),]
df<-df[order(df$GEO_ID),]



```

#### Reading shape file
```{r,echo=FALSE,warning=FALSE,message=FALSE}
map<-readShapePoly("~/admin2/SL_GN_LR_polygons.shp")

# Plot of countries
plotvar <- map$CNTR_ID
plotclr <- c("red","green","blue")
class <- as.numeric(plotvar)
colcode <- plotclr[class]

dev.new()
plot(map)
plot(map, col=colcode, add=T)
legend(-15,7 ,legend=unique(df$PARENT_COUNTRY_NAME), fill= c("red","green","blue"), cex=1, bty="n")
centroids <- coordinates(map)




# The next areas are not being considered (no cases in these areas reported)
map<-map[map@data$GEO_ID != "GNG240724",] 
map<-map[map@data$GEO_ID != "GNG240732",] 
map<-map[map@data$GEO_ID != "GNG240733",] 
map<-map[map@data$GEO_ID != "GNG240734",] 
map<-map[map@data$GEO_ID != "GNG240735",] 
map<-map[map@data$GEO_ID != "GNG240736",] 
map<-map[map@data$GEO_ID != "GNG240711",]   # Koundara
map<-map[map@data$GEO_ID != "GNG240712",]   # Dixinn
map<-map[map@data$GEO_ID != "GNG240713",]   # Kaloum
map<-map[map@data$GEO_ID != "GNG240714",]   # Matam
map<-map[map@data$GEO_ID != "GNG240715",]   # Matoto
map<-map[map@data$GEO_ID != "GNG240716",]   # Ratoma

# Order areas
map<-map[order(map@data$GEO_ID),]
centroids <- coordinates(map)

# Plot of countries
plotvar <- map$CNTR_ID
plotclr <- c("red","green","blue")
class <- as.numeric(plotvar)
colcode <- plotclr[class]

plot(map)
plot(map, col=colcode, add=T)
legend(-15,7 ,legend=unique(df$PARENT_COUNTRY_NAME), fill= c("red","green","blue"), cex=1, bty="n")
text(centroids, label=map@data$GEO_NAME,cex=0.3)


```

#### In Summary:

- Country of interest  is `r unique(df$PARENT_COUNTRY_NAM)`.

- Total number of areas is `r length(unique(df$GEO_ID))`.

- Number of weeks is `r max(df$week)-min(df$week)+1`.

- Time period of interest is from `r sort(unique(df$week))[1]` until `r tail(sort(unique(df$week)),1)[1]`.

### 2) Geographical distribution of total number of cases
```{r,echo=FALSE,warning=FALSE}
narea<-length(unique(df$GEO_ID))
cases<-df[df$Status=="CASE",]
deaths<-df[df$Status=="DEATH",]
cumcases<-unlist(sapply(1:narea,function(i) tail(cases$CountW[cases$GEO_ID==unique(df$GEO_ID)[i]],1)))
cumdeaths<-unlist(sapply(1:narea,function(i) tail(deaths$CountW[deaths$GEO_ID==unique(df$GEO_ID)[i]],1)))
cumcases2<-unlist(sapply(1:narea,function(i) max(cases$CountW[cases$GEO_ID==unique(df$GEO_ID)[i]])))
cumdeaths2<-unlist(sapply(1:narea,function(i) max(deaths$CountW[deaths$GEO_ID==unique(df$GEO_ID)[i]])))


# plot cases
plotvar <- cumcases
nclr <- 5
plotclr <- brewer.pal(nclr,"Reds")
class <- classIntervals(plotvar, nclr, style="fixed",fixedBreaks=c(0,1,10,100,1000,6000))
colcode <- findColours(class, plotclr)

plot(map)
plot(map, col=colcode, add=T)
x0<-bbox(map)[1,1]-0.4
y0<- bbox(map)[2,1]+2*bbox(map)[2,2]/(bbox(map)[2,2]-bbox(map)[2,1])
legend(x0,y0, legend=names(attr(colcode, "table")), fill=attr(colcode, "palette"), cex=1, bty="n")
centroids <- coordinates(map)

title('Cumulative Number of Cases')

# plot deaths
plotvar <- cumdeaths
nclr <- 4
plotclr <- brewer.pal(nclr,"Reds")
class <- classIntervals(plotvar, nclr, style="fixed",fixedBreaks=c(0,1,10,100,1000,2500))
colcode <- findColours(class, plotclr)

par(mar = c(3, 3, 2, 0.5), mgp = c(2, 0.5, 0), tcl = -0.3, cex.axis = 0.8,
    cex.lab = 0.8, cex.main = 1)
plot(map)
plot(map, col=colcode, add=T)
x0<-bbox(map)[1,1]-0.4
y0<- bbox(map)[2,1]+2*bbox(map)[2,2]/(bbox(map)[2,2]-bbox(map)[2,1])
legend(x0,y0, legend=names(attr(colcode, "table")), fill=attr(colcode, "palette"), cex=1, bty="n")
centroids <- coordinates(map)
title('Cumulative Number of Deaths')

```

```{r,echo=FALSE,warning=FALSE}
#### neighborhood structure to be considered in the analysis

mapqnb<-poly2nb(map,queen=T)
g<-nb2mat(mapqnb)

graph = g
```



```{r,echo=FALSE,warning=FALSE}
#### Restructure data for the analyses
df2<-merge(cases,deaths,by=c("GEO_ID","week"),all=T)
df2<-df2[,c(1,2,4,5,7,13)]
names(df2)<-c("GEO_ID","week","GEO_NAME","PARENT_COUNTRY_NAME","cumcases","cumdeaths")

region.unstruct<-unlist(sapply(1:nrow(df2),function(i) {(1:narea)[unique(df2$GEO_ID)==df2$GEO_ID[i]]}))
df2$region.unstruct<-region.unstruct
df2$region.struct<-region.unstruct


cumcases.new<-NULL
cases.new<-NULL
for (i in 1:narea)
{
  hlp<-df2$cumcases[df2$GEO_ID==unique(df2$GEO_ID)[i]]
  if (sum(!is.na(hlp))==0) {hlp<-0}  
  last<-tail(hlp,1)
  if (is.na(last)) {last<-hlp[length(hlp)-1];hlp[length(hlp)]<-last}
  if (length(hlp)>1){
    for (j in (length(hlp)-1):1)
    {
      if (is.na(hlp[j])){hlp[j]<-last}
      if (hlp[j]>last) {hlp[j]<-last}
      else {last<-hlp[j]}
    }
    hlp2<-hlp
    hlp2[2:length(hlp)]<-hlp[2:length(hlp)]-hlp[1:(length(hlp)-1)]
  }
  if (length(hlp)==1) {hlp2<-hlp}
  cumcases.new<-c(cumcases.new,hlp)
  cases.new<-c(cases.new,hlp2)
}

cumdeaths.new<-NULL
deaths.new<-NULL
for (i in 1:narea)
{
  hlp<-df2$cumdeaths[df2$GEO_ID==unique(df2$GEO_ID)[i]]
  last<-tail(hlp,1)
  if (is.na(last)) {last<-hlp[length(hlp)-1];hlp[length(hlp)]<-last}
  if (length(hlp)>1){
    for (j in (length(hlp)-1):1)
    {
      if (is.na(hlp[j])){hlp[j]<-last}
      if (hlp[j]>last) {hlp[j]<-last}
      else {last<-hlp[j]}
    }
    hlp2<-hlp
    hlp2[2:length(hlp)]<-hlp[2:length(hlp)]-hlp[1:(length(hlp)-1)]
  }
  if (length(hlp)==1) {hlp2<-hlp}
  cumdeaths.new<-c(cumdeaths.new,hlp)
  deaths.new<-c(deaths.new,hlp2)
}



df2$cumcases.new<-cumcases.new
df2$cumdeaths.new<-cumdeaths.new
df2$cases<-cases.new
df2$deaths<-deaths.new


data<-cbind(df2$week,df2$week-min(df2$week),df2$cases,df2$deaths,df2$cumcases.new,df2$cumdeaths.new,df2$region.unstruct,df2$region.struct)
colnames(data)<-c("week","time","cases","deaths","cumcases","cumdeaths","region.unstruct","region.struct")
data<-as.data.frame(data)


data$country<-1
data$country[df2$PARENT_COUNTRY_NAME=="Sierra Leone"]<-2
data$country[df2$PARENT_COUNTRY_NAME=="Guinea"]<-3
data$country[is.na(df2$PARENT_COUNTRY_NAME)]<-3
data$country2<-0
data$country2[data$country==2]<-1
data$country3<-0
data$country3[data$country==3]<-1

data<-rbind(data,
            c(20,20,0,0,221,5,42,42,2,1,0),
            c(21,21,0,0,221,5,42,42,2,1,0),
            c(20,20,0,0,327,11,43,43,2,1,0),
            c(21,21,0,0,327,11,43,43,2,1,0))


data.add<-as.data.frame(cbind(rep(seq(51,57),15),rep(seq(51,57),15),rep(0,15*7),rep(0,15*7),
                              unlist(rep(sapply(28:42,function(i){rep(max(data$cumcases[data$region.unstruct==i]),7)}))),
                              unlist(rep(sapply(28:42,function(i){rep(max(data$cumdeaths[data$region.unstruct==i]),7)}))),
                              rep(28:42,each=7),rep(28:42,each=7),rep(1,15*7),rep(0,15*7),rep(0,15*7)))
colnames(data.add)<-c("week","time","cases","deaths","cumcases","cumdeaths","region.unstruct","region.struct","country","country2","country3")

data<-rbind(data,data.add)

# Order data
data<-data[order(data$week),]
data<-data[order(data$region.unstruct),]


```

### 3)  Plot time-series of new cases per region

```{r,echo=FALSE,warning=FALSE}
par(mfrow=c(1,1),mar=c(5,3,2,2)+0.1)

plot(data$week,data$cases,xlab="",ylab="new cases", ylim=c(0,1500),type="n",xaxt="n")
axis(1, at=c(0,4,8,12,16,20,24,28,32,36,40,44,48,52,56,60),labels=c("2014-21","2014-25","2014-29","2014-33","2014-37","2014-41","2014-45","2014-49","2015-01","2015-05","2015-09","2015-13","2015-17","2015-21","2015-25","2015-29"), las=2)
for (i in 1:narea){lines(data$week[data$region.struct==i],data$cases[data$region.struct==i]) }
title('Number of new cases per week')
#abline(v=c(27,33,37),col="red")

u<-sort(unique(data$week))
plot(u,sapply(1:length(u),function(i){sum(data$cases[data$week==u[i]])}),type="l",xlab="",ylab="new cases (overall)",xaxt="n")
axis(1, at=c(0,4,8,12,16,20,24,28,32,36,40,44,48,52,56,60),labels=c("2014-21","2014-25","2014-29","2014-33","2014-37","2014-41","2014-45","2014-49","2015-01","2015-05","2015-09","2015-13","2015-17","2015-21","2015-25","2015-29"), las=2)
#abline(v=c(27,33,37),col="red")
title('Number of new cases per week (overall)')

plot(data$week,data$deaths,xlab="",ylab="new deaths", ylim=c(0,500),type="n",xaxt="n")
axis(1, at=c(0,4,8,12,16,20,24,28,32,36,40,44,48,52,56,60),labels=c("2014-21","2014-25","2014-29","2014-33","2014-37","2014-41","2014-45","2014-49","2015-01","2015-05","2015-09","2015-13","2015-17","2015-21","2015-25","2015-29"), las=2)
for (i in 1:narea){lines(data$week[data$region.struct==i],data$deaths[data$region.struct==i]) }
#abline(v=c(27,33,37),col="red")
title('Number of new deaths per week')

plot(data$week,data$cumcases,xlab="",ylab="cumulative cases", ylim=c(0,5000),type="n",xaxt="n")
axis(1, at=c(0,4,8,12,16,20,24,28,32,36,40,44,48,52,56,60),labels=c("2014-21","2014-25","2014-29","2014-33","2014-37","2014-41","2014-45","2014-49","2015-01","2015-05","2015-09","2015-13","2015-17","2015-21","2015-25","2015-29"), las=2)
for (i in 1:narea){lines(data$week[data$region.struct==i],data$cumcases[data$region.struct==i]) }
title('Number of cumulative cases per week')

plot(data$week,data$cumdeaths,xlab="",ylab="cumulative deahts", ylim=c(0,2000),type="n",xaxt="n")
axis(1, at=c(0,4,8,12,16,20,24,28,32,36,40,44,48,52,56,60),labels=c("2014-21","2014-25","2014-29","2014-33","2014-37","2014-41","2014-45","2014-49","2015-01","2015-05","2015-09","2015-13","2015-17","2015-21","2015-25","2015-29"), las=2)
for (i in 1:narea){lines(data$week[data$region.struct==i],data$cumdeaths[data$region.struct==i]) }
title('Number of cumulative deaths per week')




cumcases.mat<-matrix(rep(NA,length(map@data$GEO_ID)*length(unique(df$Week))),nrow=length(map@data$GEO_ID))
cumdeaths.mat<-matrix(rep(NA,length(map@data$GEO_ID)*length(unique(df$Week))),nrow=length(map@data$GEO_ID))
cases.mat<-matrix(rep(NA,length(map@data$GEO_ID)*length(unique(df$Week))),nrow=length(map@data$GEO_ID))
deaths.mat<-matrix(rep(NA,length(map@data$GEO_ID)*length(unique(df$Week))),nrow=length(map@data$GEO_ID))

v<-sort(unique(data$region.unstruct))
for (i in 1:length(map@data$GEO_ID))
{
  cumcases.mat[i,(data$week[data$region.unstruct==i]+1)]<-data$cumcases[data$region.unstruct==i]
  cumdeaths.mat[i,(data$week[data$region.unstruct==i]+1)]<-data$cumdeaths[data$region.unstruct==i]
  cases.mat[i,(data$week[data$region.unstruct==i]+1)]<-data$cases[data$region.unstruct==i]
  deaths.mat[i,(data$week[data$region.unstruct==i]+1)]<-data$deaths[data$region.unstruct==i]
}

rownames(cumcases.mat)<-map@data$GEO_ID
rownames(cumdeaths.mat)<-map@data$GEO_ID
rownames(cases.mat)<-map@data$GEO_ID
rownames(deaths.mat)<-map@data$GEO_ID

colnames(cumcases.mat)<-sort(unique(df$Week))
colnames(cumdeaths.mat)<-sort(unique(df$Week))
colnames(cases.mat)<-sort(unique(df$Week))
colnames(deaths.mat)<-sort(unique(df$Week))


save(cumcases.mat,file="cumcases.RData")
save(cumdeaths.mat,file="cumdeaths.RData")
save(cases.mat,file="cases.RData")
save(deaths.mat,file="deaths.RData")





```


#### 4) Data ETU
```{r,echo=FALSE,warning=FALSE}

########################## 
# Read in data


etu=read.xlsx("~/Dropbox/Christel/Onderzoek/datasets/Ebola/Data/Table_Ebola_RESPONSE_Updated_20150505_copy.xlsx",sheetIndex=1,startRow=99)


etu<-etu[,c(1,3,6,8,9,10,11,12,14,15,16,18,20,21,23,24,25,26,27,28,30)]
names(etu)<-c("etu","GEO_NAME","GEO_ID","W18","W2","W47","W46","W45","W44","W41","W38","W35","W34","W33","W31","W30","W27","W26","W25","W16","W13")

map@data$GEO_ID<-as.factor(as.character(map@data$GEO_ID))
etu_new<-matrix(rep(0,length(unique(df$GEO_ID))*length(unique(df$Week))),ncol=length(unique(df$Week)))
rownames(etu_new)<-map@data$GEO_ID
colnames(etu_new)<-sort(unique(df$Week))

etu_coord<-matrix(rep(NA,length(unique(df$GEO_ID))*2),ncol=2)
for (i in 1:nrow(etu))
{
  etu_new[levels(map@data$GEO_ID)==etu$GEO_ID[i],]<-as.matrix(etu[i,c("W16","W16","W16","W16","W25","W26","W27","W27","W27","W30","W31","W31","W33","W34","W35","W35","W35","W38","W38","W38","W41","W41","W41","W44","W45","W46","W47","W47","W47","W47","W47","W47","W47","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2","W2") ])
  etu_coord[levels(map@data$GEO_ID)==etu$GEO_ID[i],]<-coordinates(map)[levels(map@data$GEO_ID)==etu$GEO_ID[i],]
}
# OPMERKING: AREAS in rijen; WEKEN in kolommen week 2014-21 (0) tot 2015-05 (37)




etu_new[etu_new==1]<-0
etu_new[etu_new==2]<-2
etu_new[etu_new==3]<-1
etu_new[etu_new==4]<-2
etu_new[etu_new==5]<-1
etu_new[etu_new==6]<-2


save(etu_new,file="etu_update.RData")

```


#### 5) MOVIE
```{r,echo=FALSE,warning=FALSE}


# movie cumulative number of cases 

nclr <- 5
plotclr <- c("white",brewer.pal(nclr,"Reds"))
class <- classIntervals(data$cumcases,5, style="fixed",fixedBreaks=c(0,1,10,100,1000,3000,6000))
colcode <- findColours(class, plotclr)

u<-unique(sort(df$week))
v<-unique(sort(df$Week))

library(animation)

saveHTML({
  ani.options(interval = 0.5, nmax = 30)
  par(mar = c(3, 3, 2, 0.5), mgp = c(2, 0.5, 0), tcl = -0.3, cex.axis = 0.8,
      cex.lab = 0.8, cex.main = 1)
  
  for (i in 1:length(u))
  {
    var<-data$cumcases[data$week==u[i]]
    code<-rep("white",narea)
    code[data$region.unstruct[data$week==u[i]]]<-colcode[data$week==u[i]]
    plot(map)
    plot(map, col=code, add=T)
    legend("bottomleft", legend=names(attr(colcode, "table")), fill=attr(colcode, "palette"), cex=1, bty="n",
           title="cumul cases")
    centroids <- coordinates(map)
    
    
    col=c("#B38481","#7F525D")[etu_new[,i][(etu_new[,i]>0)]]
    points(etu_coord[(etu_new[,i]>0),],pch=18,col=col)
    #points(etu_coord[(etu_new[,i]>1),],pch=18,col="#B38481")
    title(c("Week" , as.character(v[i])))
    legend("bottom", legend=c("1","2"), col=c("#B38481","#7F525D"),pch=18, bty="n",title="ETU")
  }
  
}, img.name = "cumcases", imgdir="cumcases",title = "Ebola Cumulative Number of Cases",outdir=getwd())


# movie cumulative number of deaths

nclr <- 5
plotclr <- c("white",brewer.pal(nclr,"Reds"))
class <- classIntervals(data$cumdeaths,5, style="fixed",fixedBreaks=c(0,1,10,110,1100,2200,4600))
colcode <- findColours(class, plotclr)

u<-unique(sort(df$week))
v<-unique(sort(df$Week))

library(animation)

saveHTML({
  ani.options(interval = 0.5, nmax = 30)
  par(mar = c(3, 3, 2, 0.5), mgp = c(2, 0.5, 0), tcl = -0.3, cex.axis = 0.8,
      cex.lab = 0.8, cex.main = 1)
  
  for (i in 1:length(u))
  {
    var<-data$cumdeaths[data$week==u[i]]
    code<-rep("white",narea)
    code[data$region.unstruct[data$week==u[i]]]<-colcode[data$week==u[i]]
    plot(map)
    plot(map, col=code, add=T)
    legend("bottomleft", legend=names(attr(colcode, "table")), fill=attr(colcode, "palette"), cex=1, bty="n",
           title="cumul deaths")
    centroids <- coordinates(map)
    
    col=c("#B38481","#7F525D")[etu_new[,i][(etu_new[,i]>0)]]
    points(etu_coord[(etu_new[,i]>0),],pch=18,col=col)
    #points(etu_coord[(etu_new[,i]>1),],pch=18,col="#B38481")
    title(c("Week" , as.character(v[i])))
    legend("bottom", legend=c("1","2"), col=c("#B38481","#7F525D"),pch=18, bty="n",title="ETU")
  }
  
}, img.name = "cumdeaths", imgdir="cumdeaths", title = "Ebola Cumulative Number of Deaths",outdir=getwd())




```


### 4)  Non-linear Growth rate models

##### Comparison between different models:
##### -- negative binomial model
##### -- spatially structured effects (besag)
##### -- flexible time trend per area: autoregressive process (AR1) or random walk process (RW1 or RW2)
##### -- county-effect as fixed effect
```{r,echo=FALSE,warning=FALSE}
data$region.struct2<-data$region.struct

## TYPE2: The following models assume at each time the points are spatially correlated, while across the time periods, they evolve according to AR1/RW1/RW2 process. 
data$time2<-data$time+1
formula3f = cases ~  country2 + country3 +f(region.struct,model="besag", group=time2, control.group=list(model="ar1"), graph=graph)
formula3g = cases ~  country2 + country3 +f(region.struct,model="besag", group=time2, control.group=list(model="rw1"), graph=graph)
formula3h = cases ~  country2 + country3 +f(region.struct,model="besag", group=time2, control.group=list(model="rw2"), graph=graph)

# negative binomial model
fit1i <- inla(formula3f, family="nbinomial", data = data, control.predictor=list(compute=TRUE), 
              control.family=list(link="log"),control.compute=list(dic=TRUE,cpo=TRUE)) 
fit1j <- inla(formula3g, family="nbinomial", data = data, control.predictor=list(compute=TRUE), 
              control.family=list(link="log"),control.compute=list(dic=TRUE,cpo=TRUE)) 
fit1k <- inla(formula3h, family="nbinomial", data = data, control.predictor=list(compute=TRUE), 
              control.family=list(link="log"),control.compute=list(dic=TRUE,cpo=TRUE)) 

# poisson model
fit1l <- inla(formula3f, family="poisson", data = data, control.predictor=list(compute=TRUE), 
              control.family=list(link="log"),control.compute=list(dic=TRUE,cpo=TRUE)) 
fit1m <- inla(formula3g, family="poisson", data = data, control.predictor=list(compute=TRUE), 
              control.family=list(link="log"),control.compute=list(dic=TRUE,cpo=TRUE)) 
fit1n <- inla(formula3h, family="poisson", data = data, control.predictor=list(compute=TRUE), 
              control.family=list(link="log"),control.compute=list(dic=TRUE,cpo=TRUE)) 

```

##### Model comparison
```{r,echo=FALSE,warning=FALSE}
dic.models<-rbind(
  cbind("ar1+nbin Type2",fit1i$dic$dic,fit1i$dic$p.eff,-mean(log(fit1i$cpo$cpo))),
  cbind("rw1+nbin Type2",fit1j$dic$dic,fit1j$dic$p.eff,-mean(log(fit1j$cpo$cpo))),
  cbind("rw2+nbin Type2",fit1k$dic$dic,fit1k$dic$p.eff,-mean(log(fit1k$cpo$cpo))),
  cbind("ar1+poi Type2",fit1l$dic$dic,fit1l$dic$p.eff,-mean(log(fit1l$cpo$cpo))),
  cbind("rw1+poi Type2",fit1m$dic$dic,fit1m$dic$p.eff,-mean(log(fit1m$cpo$cpo))),
  cbind("rw2+poi Type2",fit1n$dic$dic,fit1n$dic$p.eff,-mean(log(fit1n$cpo$cpo)))
)
colnames(dic.models)<-c("model","DIC","pD","cpo")

dic.models

fit1<-fit1i  # model has smallest DIC, smallest pD and smallest CPO

summary(fit1)

```

> dic.models
model            DIC                    pD                     cpo               
[1,] "ar1+nbin Type2" "10660.0525381559"     "583.046623714203"     "2.93881494413852"
[2,] "rw1+nbin Type2" "11793.7802165965"     "1218.03248710857"     "13.1988760709058"
[3,] "rw2+nbin Type2" "NaN"                  "NaN"                  "23.5175269572577"
[4,] "ar1+poi Type2"  "8543.73056865342"     "1374.90931175847"     "6.16062029163567"
[5,] "rw1+poi Type2"  "1.40522744928232e+20" "70261372464116064256" "15.2452845458308"
[6,] "rw2+poi Type2"  "NaN"                  "NaN"                  "39.3093488052"  

##### Plot results growth rate model 

```{r,echo=FALSE,warning=FALSE}
### PLOT 1
plot(data$week,log(data$cases),xlab="week",ylab="log(new cases)",type="n")
for (i in 1:narea){
  lines(data$week[data$region.struct==i],log(data$cases[data$region.struct==i]),type="b",
        col="blue") 
}

title("Estimated number of new cases")
for (i in 1:length(unique(data$region.unstruct)))
{
  u<-unique(data$region.unstruct)
  lines(data$week[data$region.unstruct==u[i]],fit1$summary.linear.predictor[data$region.unstruct==u[i],4])
}


### PLOT 2
plot(data$week,log(data$cumcases),xlab="week",ylab="log(cases)",type="n")
for (i in 1:narea){
  lines(data$week[data$region.struct==i],log(data$cumcases[data$region.struct==i]),type="b",
        col="blue") 
}

title("Estimated number of cumulative cases")
for (i in 1:length(unique(data$region.unstruct)))
{
  u<-unique(data$region.unstruct)
  lines(data$week[data$region.unstruct==u[i]],log(cumsum(exp(fit1$summary.linear.predictor[data$region.unstruct==u[i],4]))))
}


### PLOT 3
u<-unique(data$region.unstruct)
i<-1
plot(min(data$week):max(data$week),fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]],type="n",ylim=c(-20,20),ylab="f(t)",xlab="week")
for (i in 1:length(unique(data$region.unstruct)))
{
  u<-unique(data$region.unstruct)
  col<-"black"
  diff<-fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]][22]-fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]][1]
  col[diff>0]<-"red"
  lines(min(data$week):max(data$week),fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]],col=col)
}
title("Estimated Time Trend (without peaks)")


```



##### Plots time trend per area
```{r,echo=FALSE,warning=FALSE} 

trend.country<-function(i)
{
  t<-sort(unique(data$week))
  u<-unique(data$region.unstruct)
  col<-"black"
  diff<-fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]][22]-fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]][1]
  col[diff>0]<-"red"
  
  select<-unique(data$week[data$region.unstruct==u[i]])-min(t)+1
  plot(t,fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]],col=col,ylab="f(t)",xlab="week",type="n",ylim=c(-9,6))
  lines(t[select],fit1$summary.random$region.struct$mean[fit1$summary.random$region.struct$ID==u[i]][select],col=col,ylab="f(t)",xlab="week",type="l",ylim=c(-9,6))
  
}

# Guinea
par(mfrow=c(2,4))
trend.country(i=1); title("Boke")
trend.country(i=2); title("Conakry")
trend.country(i=3); title("Boffa")
trend.country(i=4); title("Dabola")
trend.country(i=5); title("Dinguiraye")
trend.country(i=6); title("Faranah")
trend.country(i=7); title("Kissidougou")
trend.country(i=8); title("Kankan")
trend.country(i=9); title("Kerouane")
trend.country(i=10); title("Kouroussa")
trend.country(i=11); title("Siguiri")
trend.country(i=12); title("Coyah")

trend.country(i=13); title("Dubreka")
trend.country(i=14); title("Forecariah")
trend.country(i=15); title("Kindia")
trend.country(i=16); title("Telemele")
trend.country(i=17); title("Dalaba")
trend.country(i=18); title("Mamou")
trend.country(i=19); title("Pita")
trend.country(i=20); title("Beyla")
trend.country(i=21); title("Gueckedou")
trend.country(i=22); title("Lola")
trend.country(i=23); title("Macenta")
trend.country(i=24); title("Nzerekore")
trend.country(i=25); title("Yomou")

# Liberia
par(mfrow=c(2,4))
trend.country(i=26); title("Bomi")
trend.country(i=27); title("Bong")
trend.country(i=28); title("Gbarpolu")
trend.country(i=29); title("Grand Bassa")
trend.country(i=30); title("Grand Cape Mount")
trend.country(i=31); title("Grand Gedeh")
trend.country(i=32); title("Grand Kru")
trend.country(i=33); title("Lofa")

trend.country(i=34); title("Margibi")
trend.country(i=35); title("Maryland")
trend.country(i=36); title("Montserrado")
trend.country(i=37); title("Nimba")
trend.country(i=38); title("Rivercess")
trend.country(i=39); title("River Gee")
trend.country(i=40); title("Sinoe")

# Sierra Leone
par(mfrow=c(2,4))
trend.country(i=41); title("Western Area")
trend.country(i=42); title("Kailahun")
trend.country(i=43); title("Kenema")
trend.country(i=44); title("Kono")
trend.country(i=45); title("Bombali")
trend.country(i=46); title("Kambia")
trend.country(i=47); title("Koinadugu")
trend.country(i=48); title("Port Loko")

trend.country(i=49); title("Tonkolili")
trend.country(i=50); title("Bo")
trend.country(i=51); title("Bonthe")
trend.country(i=50); title("Moyamba")
trend.country(i=51); title("Pujehun")

```




#####  Calculation GROWTH RATE
###### plot of weekly growth rate
###### in red: 4th successive week growth<0
```{r,echo=FALSE,warning=FALSE}
##### Calculate posterior growth rate
nweeks<-length(unique(data$time))
grate<-matrix(NA,nweeks,narea)
grate.ll<-matrix(NA,nweeks,narea)
grate.ul<-matrix(NA,nweeks,narea)

# growth = u_(t+1)-u(t)
#n<-nrow(fit1m$summary.random$time)
n<-nrow(fit1$summary.random$region.struct)
BB = matrix(rep(c(rep(NA,n)),n),nrow=n)

for (i in 1:n)
{
  if (i<=narea) {BB[i,i+narea]<-1;BB[i,i]<--1}
  if (i>=(n-narea+1)) {BB[i,i]<-1;BB[i,i-narea]<--1}
  if (i>narea & i<(n-narea+1)){BB[i,i+narea]<-1/2;BB[i,i-narea]<--1/2}
  if (i>(2*narea) & i<(n-(2*narea)+1))
  {BB[i,i+2*narea]<--1/12;BB[i,i+narea]<-8/12;BB[i,i-narea]<--8/12;BB[i,i-2*narea]<-1/12}
  if (i>(3*narea) & i<(n-(3*narea)+1))
  {BB[i,i+3*narea]<-1/60;BB[i,i+2*narea]<--3/20;BB[i,i+narea]<-3/4;
   BB[i,i-narea]<--3/4;BB[i,i-2*narea]<-3/20;BB[i,i-3*narea]<--1/60}
  if (i>(4*narea) & i<(n-(4*narea)+1))
  {BB[i,i+4*narea]<--1/280;BB[i,i+3*narea]<-4/105;BB[i,i+2*narea]<--1/5;BB[i,i+narea]<-4/5;
   BB[i,i-narea]<--4/5;BB[i,i-2*narea]<-1/5;BB[i,i-3*narea]<--4/105;BB[i,i-4*narea]<-1/280}
}


lc = inla.make.lincombs(region.struct=BB)
names(lc) = "lc"

formula = cases ~  country2 + country3 +f(region.struct,model="besag", group=time2, control.group=list(model="ar1"), graph=graph)


fit1 <- inla(formula, family="nbinomial", data = data, control.predictor=list(compute=TRUE), 
             control.family=list(link="log"),lincomb =lc,control.compute=list(dic=TRUE),
             control.inla= list(lincomb.derived.only=TRUE,verbose=T)) 

grate<-fit1$summary.lincomb.derived$mean
grate.ll<-fit1$summary.lincomb.derived[,4]
grate.ul<-fit1$summary.lincomb.derived[,6]


grate<-matrix(grate,ncol=narea,byrow=T)
grate.ll<-matrix(grate.ll,ncol=narea,byrow=T)
grate.ul<-matrix(grate.ul,ncol=narea,byrow=T)


# plot of weekly growth rate
# in red: 4th successive week growth<0
grate.new<-grate*NA
par(mfrow=c(1,1))
t<-sort(unique(data$week))
plot(t,grate[,1],type="n",ylim=c(-2,2),xlab="week",ylab="growth rate")
for (i in 1:narea)
{
  lines(t,grate[,i])
  col.red<-rep("black",(nweeks-1))
  col.red[4:length(t)][(grate[4:length(t),i]<0)&(grate[3:(length(t)-1),i]<0)&(grate[2:(length(t)-2),i]<0)&(grate[1:(length(t)-3),i]<0)]<-"red"
  
  points(t,grate[,i],col=col.red)
  
  select<-unique(data$week[data$region.unstruct==u[i]])-min(t)+1
  grate.new[select,i]<-grate[select,i]
  
}
title("Weekly Growth Rate")



colnames(grate.new)<-unique(df$GEO_ID)
rownames(grate.new)<-sort(unique(df$Week))

save(grate.new,file="grate.RData")


```


##### Plots growth rate per area
```{r,echo=FALSE,warning=FALSE} 

growth.country<-function(i)
{
  t<-sort(unique(data$week))
  u<-unique(data$region.unstruct)
  col.red<-rep("black",(nweeks-1))
  #col.red[4:length(t)][(grate[4:length(t),i]<0)&(grate[3:(length(t)-1),i]<0)&(grate[2:(length(t)-2),i]<0)&(grate[1:(length(t)-3),i]<0)]<-"red"
  col.red[grate.ul[,i]<0]<-"blue"
  col.red[grate.ll[,i]>0]<-"red"
  select<-unique(data$week[data$region.unstruct==u[i]])-min(t)+1
  plot(t,grate[,i],xlab="week",ylab="growth rate",ylim=c(-2,2),type="n")
  lines(t[select],grate[select,i],xlab="week",ylab="growth rate",ylim=c(-2,2),type="l")
  points(t[select],grate[select,i],col=col.red[select])
  lines(t,rep(0,length(t)),lty=16)
}

#df2$GEO_NAME.x[df2$region.unstruct==51]

# Guinea
par(mfrow=c(2,4))
growth.country(i=1); title("Boke")
growth.country(i=2); title("Conakry")
growth.country(i=3); title("Boffa")
growth.country(i=4); title("Dabola")
growth.country(i=5); title("Dinguiraye")
growth.country(i=6); title("Faranah")
growth.country(i=7); title("Kissidougou")
growth.country(i=8); title("Kankan")
growth.country(i=9); title("Kerouane")
growth.country(i=10); title("Kouroussa")
growth.country(i=11); title("Siguiri")
growth.country(i=12); title("Coyah")

growth.country(i=13); title("Dubreka")
growth.country(i=14); title("Forecariah")
growth.country(i=15); title("Kindia")
growth.country(i=16); title("Telemele")
growth.country(i=17); title("Dalaba")
growth.country(i=18); title("Mamou")
growth.country(i=19); title("Pita")
growth.country(i=20); title("Beyla")
growth.country(i=21); title("Gueckedou")
growth.country(i=22); title("Lola")
growth.country(i=23); title("Macenta")
growth.country(i=24); title("Nzerekore")
growth.country(i=25); title("Yomou")

# Liberia
par(mfrow=c(2,4))
growth.country(i=26); title("Bomi")
growth.country(i=27); title("Bong")
growth.country(i=28); title("Gbarpolu")
growth.country(i=29); title("Grand Bassa")
growth.country(i=30); title("Grand Cape Mount")
growth.country(i=31); title("Grand Gedeh")
growth.country(i=32); title("Grand Kru")
growth.country(i=33); title("Lofa")

growth.country(i=34); title("Margibi")
growth.country(i=35); title("Maryland")
growth.country(i=36); title("Montserrado")
growth.country(i=37); title("Nimba")
growth.country(i=38); title("Rivercess")
growth.country(i=39); title("River Gee")
growth.country(i=40); title("Sinoe")

# Sierra Leone
par(mfrow=c(2,4))
growth.country(i=41); title("Western Area")
growth.country(i=42); title("Kailahun")
growth.country(i=43); title("Kenema")
growth.country(i=44); title("Kono")
growth.country(i=45); title("Bombali")
growth.country(i=46); title("Kambia")
growth.country(i=47); title("Koinadugu")
growth.country(i=48); title("Port Loko")

growth.country(i=49); title("Tonkolili")
growth.country(i=50); title("Bo")
growth.country(i=51); title("Bonthe")
growth.country(i=50); title("Moyamba")
growth.country(i=51); title("Pujehun")




u<-unique(sort(df$week))
v<-unique(sort(df$Week))

library(animation)

saveHTML({
  ani.options(interval = 0.5, nmax = 30)
  par(mar = c(3, 3, 2, 0.5), mgp = c(2, 0.5, 0), tcl = -0.3, cex.axis = 0.8,
      cex.lab = 0.8, cex.main = 1)
  
  for (i in 1:length(u))
  {
    var<-grate[i,]
    nclr <- 6
    plotclr <- rev(brewer.pal(nclr,"RdBu"))
    class <- classIntervals(var,6, style="fixed",fixedBreaks=c(-2,-1,-0.5,0,0.5,1,2))
    colcode <- findColours(class, plotclr)
    
    colcode[-c(data$region.unstruct[data$week==u[i]])]<-"white"
    
    plot(map)
    plot(map, col=colcode, add=T)
    legend("bottomleft", legend=names(attr(colcode, "table")), fill=attr(colcode, "palette"), cex=1, bty="n",
           title="growth rate")
    centroids <- coordinates(map)
    
    col=c("#B38481","#7F525D")[etu_new[,i][(etu_new[,i]>0)]]
    points(etu_coord[(etu_new[,i]>0),],pch=18,col=col)
    title(c("Week" , as.character(v[i])))
    legend("bottom", legend=c("1","2"), col=c("#B38481","#7F525D"),pch=18, bty="n",title="ETU")
    
  }
  
}, img.name = "growth", imgdir="growth", title = "Ebola Growth Rate",outdir=getwd())



```

##### Plots heat map: growth rates and interventions
```{r,echo=FALSE,warning=FALSE} 

# estimated growth rates, numbers and and ETU's
cumcases = cumcases.mat; cumdeaths = cumdeaths.mat
grate = grate.new; etu = etu_new

# linking ID to region name
cn = colnames(grate)
region.names = rep(NA,length(cn))
for (i in 1:length(cn)){
  region.names[i] = as.character(df$GEO_NAME[df$GEO_ID==cn[i]])[1]
}
colnames(grate)=region.names

cn = rownames(etu)
region.names = rep(NA,length(cn))
for (i in 1:length(cn)){
  region.names[i] = as.character(df$GEO_NAME[df$GEO_ID==cn[i]])[1]
}
rownames(etu)=region.names
etu = t(etu)

cn = rownames(cumcases)
region.names = rep(NA,length(cn))
for (i in 1:length(cn)){
  region.names[i] = as.character(df$GEO_NAME[df$GEO_ID==cn[i]])[1]
}
rownames(cumcases)=region.names
cumcases = t(cumcases); l.cumcases = log(cumcases+1) 

cn = rownames(cumdeaths)
region.names = rep(NA,length(cn))
for (i in 1:length(cn)){
  region.names[i] = as.character(df$GEO_NAME[df$GEO_ID==cn[i]])[1]
}
rownames(cumdeaths)=region.names
cumdeaths = t(cumdeaths); l.cumdeaths = log(cumdeaths+1) 

# matching grate and etu
etu.grate = matrix(NA,dim(etu)[1],dim(etu)[2]); rownames(etu.grate)=rownames(etu); colnames(etu.grate)=colnames(etu)
for (j in 1:dim(grate)[2]){
  etu.grate[,j] = etu[,colnames(etu)==colnames(grate)[j]]
  colnames(etu.grate)[j] = colnames(grate)[j]
}

# plot

x = which(etu.grate!=0,arr.ind=T)[,1]; names(x)=NULL
y = which(etu.grate!=0,arr.ind=T)[,2]; names(y)= NULL
z = c(etu.grate[etu.grate!=0])
z.col = z; z.col[z.col==1]="#B38481"; z.col[z.col==2]="#7F525D";

col = c("#0571B0","#75B4D4","#CEE3ED","#F5D6C8","#EB836E","#CA0020")
int = c(-2,-1,-0.5,0,0.5,1,2)
windows(record=TRUE, width=7.5, height=7)
levelplot(grate,col.regions=col,at=int,scales=list(x=list(rot=45,cex=.5),y=list(cex=.5)),
          ylab=list(label="Region",cex=1),xlab=list(label="Week",cex=1),
          colorkey=list(labels=list(at=int)),
          panel=function(...) {
            panel.levelplot(...)
            panel.points(x=x, y=y, type="p",pch=21,fill=z.col,col=z.col,cex=0.7)
          }) 
```
